---
title: "dplyr_test"
author: "Claudio Peroni"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# _dplyr_ test

## Data import
First we require the package of interest
```{r}
require(dplyr)
```

Then, we import a dataset from the UCI archive (a file in _csv_ format)
```{r}
# Data file
file <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"
```

We create the vector of the names of the columns of the data table we are about to create
```{r}
# Some sensible variable names
df_names <- c("age", "wrkclass", "fnlweight", "education_lvl", "edu_score",
 "marital_status", "occupation", "relationship", "ethnic", "gender",
 "cap_gain", "cap_loss", "hrs_wk", "nationality", "income")
```


Since the file is in _csv_ format, we have to import the data as a table, and we use as column names the one we have defined in the step above
```{r}
# Import the data
df <- read.csv(file, header = F,
 sep = ",",
 na.strings = c(" ?", " ", ""),
 row.names = NULL,
 col.names = df_names)
```

Finally, we can see the result
```{r}
head(df)  # Shows the first rows of the data table
```

## Application of some functions (and piping -- %>%)

First we can check how many duplicate entries are present in our data set
```{r}
total_row_n = df %>% nrow()

# Check number of rows without duplicates
unique_row_n = df %>% distinct() %>% nrow()

duplicate_row_n = total_row_n - unique_row_n
duplicate_row_n
```

We can then create a new, clean, data table by removing those `{r duplicate_row_n}` rows
```{r}
# Drop duplicate rows and assign to new dataframe object
df_clean <- df %>% distinct()
```






