---
title: "Lab1: Demo RMarkdown & EDA"
subtitle: "MAT43 Statistical Machine Learning"
author: "Silvia Montagna (edited by Claudio Peroni)"
date: "03/06/2018"
output:
  html_document: default
  pdf_document: 
    keep_tex: yes
---

```{r setup, include = FALSE}
# These are general options that apply to all code chunks (but local options overwrite the general rules)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

#  Exercise 8 from ISLR Chapter 2

The following illustrate commands for exploring this exercise using R and various packages for the `College` data. The `College` dataset contains a number of variables for 777 different universities and colleges in the US.

#### Note
On macOs, to run the next chunk the shortcut is different from what is displayed under the _run_ menu:
```{r eval=FALSE}
cmd+alt+n
```

### Libraries

Try to load the `ISLR` library
```{r}
library(ISLR)
```

If it is not available you will need to install the  library from CRAN. Click on  *Packages*  then *Install*.  Enter the package name then click on the Install button.

You can also install from the console/command line using
`install.packages("ISLR")`.

Ready?

###  Getting the College data

Next we will need to load the dataset.  This is part of the library so we will not need to read it in using `read.csv` but rather we will use the `data` function to load it from the library.

```{r data}
data(College)
```

This loads the dataframe `College`.  Note you can always see the content of any `R` object by simply typing its name, e.g. by typing `College` in the `R` console.

```{r echo=FALSE}
College
```


For information about the variables, read the text or enter
```{r}
help(College)
```

The info will appear in the `help` tab.

To explore the data, you can use the command `View(College)`. This will open a new tab, where you may scroll left and right to look at the rows and columns.

```{r}
View(College)
```

In the `View` you should see that the first column is the College/University name. These can be extracted using `rownames(College)`.  Let's print out the first 5

```{r}
rownames(College)[1:5]
```

### Summary
We can pull up basic information about the variables using the `summary()` function

```{r}
summary(College)
```

We can also display the type of variable of each column, applying the funciton `class()` with `sapply()`
```{r}
sapply(College, class)
```

### Data dimension

How many observations and variables are in the dataframe?


```{r dim, echo=TRUE}
d = dim(College)
d
```

Suppose we want to refer to those numbers in the text.  We can extract them using n = `r d[1]` and d = `r d[2]`. Look at the Rmd code to see how we extracted them. To extract themi in R:

```{r}
n = d[1]
n
r = d[2]
r
```


### Scatter plot matrices

This `base` `R` version of scatter plot matrices is obtained using the `pairs` function to plot all variables versus each other.  We can use subsetting of columns of the dataframe to look at the first 5 columns.  

```{r, fig.width=8, fig.height=8}
pairs(College[, 1:5])
```

We can also look at this using the `ggpairs` function. Install the library `GGally` if it is not available (and any dependent libraries) and load it.

```{r ggpairs, warning=FALSE, message=FALSE, fig.width=8, fig.height=8}
library(GGally)
ggpairs(College, columns= c(1,3:5, 2), title = "ggpairs of the first 5 columns of College", showStrips = TRUE, axisLabels = TRUE)
```
_*TODO*: Try to better display the scale in each plot_

The last variable `Apps` is our response. This is the reason why we postponed it to the last column. It indicates the number of applications received.

The `ggpairs` function realizes that the variable `Private` is categorial and plots side by side histograms.  The density plots are also useful for seeing the skewness in the marginal distributions.

What other features do these plots indicate?

### New variables

Let's create a new variable `Elite` by binning the `Top10perc` variable. We are going to divide universities into two groups based on whether or not the proportion of students coming from the top 10% of their high school classes exceeds 50%.  We will use the library `dplyr` to illustrate some of the possible transformations and the idea of pipes, which are quite powerful once you get the hang of them!

```{r}
library(dplyr)
College = College %>% 
  mutate(Elite = factor(Top10perc > 50)) %>%
  mutate(Elite = recode(Elite, 'TRUE' = "Yes", 'FALSE' = "No"))
College
```

_What is the above doing?_    Document the code here.

Compare to the base `R` code:

```{r, evaluate=FALSE}
Elite = rep("No", nrow(College))
Elite[College$Top10perc > 50] = "Yes"
Elite = as.factor(Elite)
college = data.frame(College, Elite)
```

_How many Elite Universities are there?_

```{r}
summary(College$Elite)
```

### Side-by-side boxplots

Let's plot the variable `Outstate` versus `Elite` using side-by-side boxplots.  Using `base R` we would enter:

```{r, fig.width = 5, fig.height = 4, fig.align = "center"}
boxplot(Outstate ~ Elite, data = College, ylab = "Outstate", xlab = "Elite")
title("Distribution of Out of State Tuition")
```

Now for the `ggplot` version:

```{r, fig.width=6}
library(ggplot2)
my.bp <- ggplot(data = College, aes(y = Outstate, x = Elite)) # Creates boxplots
my.bp <- my.bp + geom_boxplot() # Adds color
my.bp <- my.bp + ggtitle("Distribution of Out of State Tuition") # Adds a title
my.bp <- my.bp +  ylab("Outstate") + xlab("Elite") # Adds lables for axes
my.bp # displays the boxplots
```

### Conditional Plots

Let's look at the distribution of out of state tuition (`Outstate`) versus Elite status for Private versus Public universities using _conditional plots_

```{r, fig.width = 8}
coplot(Outstate ~ Elite | Private, data = College)
```

##  ggplot conditional plot

```{r, fig.width = 6, echo = TRUE, cache = TRUE, fig.align = "center"}
library(ggplot2)
ggplot(College, aes(x = Elite, y = Outstate, 
                    group = Private, 
                    colour = Private)) +
   geom_point() + facet_grid(.~Private) + ggtitle("Distribution of Out of State Tuition") 
```

## Next Steps

Update this document and explore the other variables thinking about the objective of predicting `Apps`. Document what you discover thinking about models to predict `Apps`.

#  On your own: the `Auto` dataset

The `Auto` dataset is a data frame with 392 observations on the following 9 variables:

* `mpg` : miles per gallon
* `cylinders`: number of cylinders between 4 and 8
* `displacement`: engine displacement (cu. inches)
* `horsepower`: engine horsepower
* `weight`: vehicle weight (lbs.)
* `acceleration`: time to accelerate from 0 to 60 mph (sec.)
* `year`: model year (modulo 100)
* `origin`: origin of car (1. American, 2. European, 3. Japanese)
* `name`: vehicle name

This dataset was taken from the StatLib library which is maintained at Carnegie Mellon University. The dataset was used in the 1983 American Statistical Association Exposition.
The dataset is available with the ISLR library.

Load the data and answer the following questions adding your code in the code chunks. 

1. Create a summary of the data. How many variables have missing data?

```{r}
library(ISLR)
data(Auto)
summary(Auto)
summary(is.na(Auto))
```



2. Which of the predictors are quantitative, and which are qualitative?

```{r}
help(Auto)
typeAuto = sapply(Auto, class) #Automatic type detection
typeAuto
```



3. What is the range of each quantitative predictor? You can answer this using the `range()` function. Create a table with variable name, min, max with one row per variable. `kable` from the package `knitr` can display tables nicely.

```{r}
library(knitr)

table = t(sapply(Auto[typeAuto=="numeric"], range)) # Creates min-max table only for numerical values
kable(table, col.names = c("min", "max"))
```

4. What is the mean and standard deviation of each quantitative predictor? _Format nicely in a table as above_

```{r}

```

5. Now remove the 10th through 85th observations (try this with `filter` from the `dplyr` package). What is the range, mean, and standard deviation of each predictor in the subset of the data that remains? _Again, present the output as a nicely formatted table_

```{r}

```

6. Investigate the predictors graphically, using scatterplot matrices (`ggpairs`) and other tools of your choice. Create some plots highlighting the relationships among the predictors. Comment on your findings. _Try adding a caption to your figure_

```{r, fig.align = "center"}

```


7. Suppose that we wish to predict gas mileage (`mpg`) on the basis of the other variables using regression. Do your plots suggest that any of the other variables might be useful in predicting mpg using linear regression? Justify your answer.

